---
title: "Project Proposal"
author: "Jenni 23988835\\ Max Yin 12580312\\ Yingming Sha 39664552"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Team Information

**Team members:** Max Yin, Jenni Tong, Yingming Sha

**Public repository:**\
<https://github.com/ymyms/stat-405-project>

The repository contains commits from all team members and will be used to store data, code and the report.

# Candidate Datasets

## Main Dataset: NBA Players Dataset

**URL:**\
<https://www.kaggle.com/datasets/justinas/nba-players-data>

This dataset contains information on players who have appeared on NBA team rosters over more than two decades. Each observation corresponds to a player-season entry and includes demographic variables such as age, height, weight, and place of birth, along with biographical details like the team played for, draft year, and draft round. The dataset also includes basic box score statistics such as games played and average numbers of points, rebounds, and assists. Together, these variables provide both physical characteristics of players and measures of their performance.

```{r}
nba <- read.csv("all_seasons.csv")
head(nba)
```

This dataset is suitable for our project because it includes both player attributes and performance statistics that can be analyzed using Bayesian. In particular, variables such as height allow us to study the relationship between players’ physical characteristics. Since the dataset contains many observations across multiple seasons, it provides enough data to estimate posterior distributions and compare different Bayesian modeling approaches.

## Backup Dataset: Olympic Athletes Dataset

**URL:**\
<https://www.kaggle.com/datasets/abdullahmeo/120-years-of-olympic-athletes-dataset>

This dataset contains information on athletes who participated in the modern Olympic Games from Athens 1896 to Rio 2016, covering more than 120 years of Olympic history. The dataset includes 271,116 observations and 15 variables, where each row corresponds to an athlete competing in a specific Olympic event. The variables include demographic characteristics such as sex, age, height, and weight, as well as information about the athlete’s team, country (NOC code), sport, and event. The dataset also records the year and location of the Olympic Games, along with medal outcomes (Gold, Silver, Bronze, or no medal). Overall, the dataset provides a broad overview of athlete characteristics and participation across many sports and countries over time.

```{r}
olympic <- read.csv("athlete_events.csv")
head(olympic)
```

This dataset serves as a suitable backup for our project because it also contains physical attributes such as height along with measurable outcomes such as medal results. These variables allow similar Bayesian analyses to those proposed for the NBA dataset, such as modeling relationships between physical characteristics and athletic performance. In addition, the large number of observations across many years provides sufficient data for estimating posterior distributions and comparing Bayesian models if we choose to use this dataset.

# Scientific Questions: 

Our main goal is to study height advantages in NBA draft round in the recent 20 years, in the aim to assist our prediction of rookie player's future draft outcomes. 



```{r}
recent_10 <- nba[nba$draft_year>=2006 & !(nba$draft_round== "Undrafted"),]

rookie_data <- recent_10 |>
  dplyr::arrange(player_name, season) |>
  dplyr::distinct(player_name, .keep_all = TRUE)

quantile(rookie_data$player_height,c(0.25,0.5,0.75))

```

Knowing that the upper quantile is 208.28cm, we wonder whether player that are higher than upper quantile have advantages in draft round. 

Our primary scientific question is:

**What is the probability that a player was drafted in the first round given that their height exceeds 208.28 cm?**




A secondary question we may investigate is:

**What is the probability that a player was drafted in the first round given that their height is below 195.58 cm?**

These questions are well suited to Bayesian methods because they allow us to quantify uncertainty, incorporate prior information, and compare posterior distributions under different model assumptions.

# Proposed Methodology

## Data Analysis

- Perform Exploratory Data Analysis to visualize, clean ,and explore potential relationship between variables.

- Assign prior and likelihood distribution based on observations, visualizations, or knowledge from prior studies.

## Posterior Computation Plan

We plan to use at least two methods for posterior computation:

1.  SNIS
2.  MCMC

This will allow us to compare posterior approximations and assess their agreement. We will discuss if the results support one another, and why such agreement or disagreement happens.

## Model Checking and Validation

We plan to evaluate our Bayesian models using:

-   Knowledge from prior research studies
-   Assess credible interval and check if it is approximately calibrated
-   Visualize the posterior distribution


# Team Contribution Plan

To ensure all team members contribute roughly equally, we plan the following division of work:

-   **Max Yin:** 

    Exploratory Data Analysis
    
    Discussion and Limitations
    
    Compare the results from the two methods
    
-   **Jenni Tong:** 

    Constructing Prior and likelihood Distribution
    
    Implement MCMC method
    
    Analyze posterior distribution
    
-   **Yingming Sha:** 

    Constructing Posterior Distribution
    
    Implement SNIS method
    
    Interpret posterior result

All team members will contribute to background introduction, literature review, reference.